---
title: <font size = "6">Demo of microbial fcm analysis pipeline</font>
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: flatly
    toc: yes
    number_sections: true
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
editor_options:
  chunk_output_type: console
---

<style type="text/css">
h1.title {
  font-size: 38px;
  color: black;
  text-align: center;
}

</style>

```{r setup-1, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      cache = FALSE, 
                      include = TRUE, 
                      collapse = FALSE, 
                      dependson = NULL, 
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.path="./Figures/",  # Set the figure options
                      fig.align = "center"
                      )
options(knitr.duplicate.label = "allow")
```

# Libraries

```{r, include = TRUE, warning = FALSE, message = FALSE}
library("Phenoflow")
library("plyr")
library("dplyr")
library("ggplot2")
library("flowAI")
library("scales")
library("plyr")
library("cowplot")
library("ggcyto")
library("RColorBrewer")
library("grid")
library("tidyr")
library("flowFP")
library("FlowSOM")
library("readxl")
# for flowEMMI
library("Rcpp")
library("RcppEigen")
library("mixtools")
library("gtools")
library("flowCore")
library("flowViz")
library("randomcoloR")
sourceCpp("ext/flowEMMi.cpp")

# Set seed for reproducible analysis
set.seed(777)

# For avoiding verbose console output
run_quiet <- function (x) {
  sink(tempfile())
  on.exit(sink())
  invisible(force(x))
}

# Import metadata
metadata <- readxl::read_excel("./metadata.xlsx")
metadata$Sample_names <- gsub(".5", "_5", metadata$Sample_names, fixed = TRUE)
metadata$Sample_names <- gsub(".fcs", "_QC.fcs", metadata$Sample_names, fixed = TRUE)
metadata$Treatment <- factor(metadata$Treatment,
                             levels = c("Adaptation of AMC",
                                        "AMC after temperature disturbance",
                                        "AMC after pH disturbance",
                                        "Adaptation of CMC",
                                        "CMC after temperature disturbance",
                                        "CMC after pH disturbance"))
```

# Preprocessing

## Import data 

Data set from [Liu et al (2017).](https://msphere.asm.org/content/3/1/e00564-17)

```{r, include = TRUE, cache = TRUE, cache.lazy = FALSE}
# Import data
flowData <- read.flowSet(path = "data/zishu_2018/")
param <- c("FS Log", "SS Log","FL 4 Log")
sampleNames(flowData) <- gsub(".5", "_5", sampleNames(flowData), fixed = TRUE)
```

## Transform data 

```{r, include = TRUE, cache = TRUE, cache.lazy = FALSE}
# Transform data
flowData_transformed_log <- transform(flowData,`FS Log`=log10(`FS Log`),
                                   `SS Log`=log10(`SS Log`),
                                   `FL 4 Log`=log10(`FL 4 Log`))
flowData_transformed_asinh <- transform(flowData,`FS Log`=asinh(`FS Log`),
                                   `SS Log`=asinh(`SS Log`),
                                   `FL 4 Log`=asinh(`FL 4 Log`))
```

## Denoise data - step 1 {.tabset}

### Untransformed data 

```{r, include=TRUE}
r_sam <- sample(1:length(flowData), 6)

#  scatter plot
p_scatter1 <- ggcyto::ggcyto(flowData[r_sam], aes(x = `FS Log`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Forward scatter (a.u.)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))
  # coord_cartesian(xlim = c(0,6), ylim = c(0,6))

print(p_scatter1)

```

### Log-transformed data 

```{r, include=TRUE}
#  scatter plot
p_scatter2 <- ggcyto::ggcyto(flowData_transformed_log[r_sam], aes(x = `FS Log`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Forward scatter (a.u.)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))
  # coord_cartesian(xlim = c(0,6), ylim = c(0,6))

print(p_scatter2)
```

### asinh-transformed data 

```{r, include=TRUE}
#  scatter plot
p_scatter3 <- ggcyto::ggcyto(flowData_transformed_asinh[r_sam], aes(x = `FS Log`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Forward scatter (a.u.)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))
  # coord_cartesian(xlim = c(0,6), ylim = c(0,6))

print(p_scatter3)
```

## Denoise data - step 2 {.tabset}

### Without gate

```{r, include=TRUE}
#  scatter plot
p_scatter4 <- ggcyto::ggcyto(flowData_transformed_asinh[r_sam], aes(x = `FS Log`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Forward scatter (a.u.)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))
  # coord_cartesian(xlim = c(0,6), ylim = c(0,6))

print(p_scatter4)
```

### With gate

```{r, include=TRUE}
# Create a PolygonGate for denoising the dataset
polycut <- matrix(c(1.5, 1.5, 4, 4, 9, 9, 6.75, 6.75, 6.25, 6.25,
  2.25, 6, 6, 8.75, 8.75, 2.25, 2.25, 2.75, 2.75, 2.25),
     ncol = 2,
     nrow = 10)
colnames(polycut) <- c("FS Log", "FL 4 Log")
polyGate <- polygonGate(.gate = polycut, filterId = "Cell population")

#  scatter plot
p_scatter5 <- ggcyto::ggcyto(flowData_transformed_asinh[r_sam], aes(x = `FS Log`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Forward scatter (a.u.)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))+
  # coord_cartesian(xlim = c(0,6), ylim = c(0,6))+
  geom_gate(polyGate, col = "#CB0001", fill = "#ffa401", alpha = 0.8, size = 1)
  
print(p_scatter5)

# Retain only events in gate
flowData_transformed_asinh <- Subset(flowData_transformed_asinh, polyGate)
```

## Denoise data - step 3

Doublet/clump removal not possible for this dataset due to lack of combined
-A/-H parameters.

## Denoise data - step 4 {.tabset}

### Before flowAI

```{r, include=TRUE}
#  scatter plot
p_scatter6 <- ggcyto::ggcyto(flowData_transformed_asinh[r_sam], aes(x = `TIME`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Time (ms)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))

print(p_scatter6)
```

```{r, include=TRUE, message =  FALSE, echo = FALSE, warning = FALSE, cache = TRUE, cache.lazy = FALSE}
ChExcludeFM <- 
  colnames(flowData_transformed_asinh)[!colnames(flowData_transformed_asinh) %in% param]

# Use flowAI for removing anomalies from the data
# flowData_transformed_asinh_dn <- run_quiet(flow_auto_qc(flowData_transformed_asinh, 
#   output = 1,
#   folder_results = "./data/QC_flowAI",
#   ChExcludeFM = ChExcludeFM,
#   timeCh = "TIME",
#   second_fractionFR = "timestep"
#   )
# )

flowData_transformed_asinh_dn <- read.flowSet(path = "./data/QC_flowAI", pattern = "_QC.fcs")
```

### After flowAI 

```{r, include=TRUE}
#  scatter plot
p_scatter7 <- ggcyto::ggcyto(flowData_transformed_asinh_dn[r_sam], aes(x = `TIME`, y = `FL 4 Log`)) + 
  geom_hex(bins = 300) +
  theme_bw()+
  labs(x = "Time (ms)", y = "DAPI (a.u.)")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.background = element_rect(colour="white", fill="white"),
        panel.border = element_rect(colour = "white"))

print(p_scatter7)
```

### QC stats

```{r, include=TRUE}
QC_stats <- read.delim("./data/QC_flowAI/QCmini.txt")

p_qc1 <- QC_stats %>%
  arrange(-X..anomalies) %>% 
  mutate(Name.file = factor(Name.file, levels = unique(Name.file))) %>% 
  ggplot(., aes(y = X..anomalies, x = Name.file))+
  geom_point(shape = 21, size = 3, fill = "#ffa401", alpha = 0.5)+
  theme_bw()+
  theme(axis.text = element_text(size = 12),
    axis.text.x = element_blank(),
        axis.title = element_text(size = 12))+
  labs(y = "% anomalies",  x = "Ranked samples")+
  geom_hline(yintercept = 5, linetype = 2)

p_qc2 <- QC_stats %>% 
  ggplot(., aes(x = X..anomalies))+
  geom_histogram(fill = "#ffa401", alpha = 0.5, col = "black")+
  theme_bw()+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12))+
  labs(x = "% anomalies",  y = "Number of samples")+
  geom_vline(xintercept = median(QC_stats$X..anomalies), linetype = 2)+
  scale_x_continuous(breaks = seq(from = 0, to = 100, by = 10), limits = c(-5,100))

cowplot::plot_grid(p_qc1, p_qc2, align = "hv", ncol = 1)
```

# Cell density measurement

Cell concentration estimates cannot be easily estimated for this dataset due to 
a lack of volumetric measurements. We refer the reader to the original manuscript.

```{r, include=TRUE}

```

# Cell population identification {.tabset}

## FlowSOM - fig1
```{r, include=TRUE, cache = TRUE, cache.lazy = TRUE, dpi = 500}
fSOM <- FlowSOM(
  input = flowData_transformed_asinh_dn[1:3],
  # Input options:
  compensate = FALSE,
  transform = FALSE,
  scale = FALSE,
  # SOM options:
  colsToUse = param,
  xdim = 7,
  ydim = 7,
  nClus = 10
  )

PlotStars(fSOM[[1]], backgroundValues = as.factor(fSOM[[2]]))
```

## FlowSOM - fig2
```{r, include=TRUE, dpi = 500}
PlotMarker(fSOM[[1]], param[3])
```

# Cytometric fingerprinting {.tabset}

## PhenoGMM

```{r, include=TRUE, dpi = 500}
fp_gmm <-
  PhenoGMM(
    flowData_transformed_asinh_dn,
    param = param,
    downsample = 1e2,
    auto_nG = TRUE,
    nG = 32,
    diagnostic_plot = TRUE
  )

# Calculate diversity indices
div_gmm <- Diversity_gmm(fp_gmm, R = 10)

# Merge with metadata
div_gmm <- left_join(div_gmm, metadata, by = "Sample_names")
```

```{r, include=TRUE, dpi = 500, fig.width = 10, fig.height = 5}
# Plot results
div_gmm %>% 
  ggplot(., aes(x = Timepoint,  y = D2, fill = Treatment))+
  geom_point(shape = 21, size = 4)+
  geom_line()+
  facet_wrap(.~Treatment, scales = "free_x", ncol = 3)+
  scale_fill_brewer("", palette = "Accent")+
  theme_bw()+
  theme(strip.text = element_text(size = 8))+
  labs(y = "Hill diversity index (D2)")

```

## PhenoFlow

```{r, include=TRUE, dpi = 500}
# Summary of max intensity values across data set
summary <-
  fsApply(
    x = flowData_transformed_asinh_dn,
    FUN = function(x)
      apply(x, 2, max),
    use.exprs = TRUE
  )
maxval <- max(summary[, "FL 4 Log"])

# Normalize intensities to [0,1] range for density estimation
mytrans <- function(x)
  x / maxval
flowData_transformed_asinh_dn_trans <-
  transform(
    flowData_transformed_asinh_dn,
    `FS Log` = mytrans(`FS Log`),
    `SS Log` = mytrans(`SS Log`),
    `FL 4 Log` = mytrans(`FL 4 Log`)
  )

# Estimate kernel densities across grids
fp_grid <-
  flowBasis(
    FCS_resample(flowData_transformed_asinh_dn_trans, 1e3),
    param = param,
    nbin = 128
  )

# Perform ordination analysis
beta_fp <- beta_div_fcm(x = fp_grid, ord.type = "PCoA")
beta_fp <- data.frame(Sample_names = rownames(beta_fp$points), beta_fp$points)
beta_fp <- left_join(beta_fp, metadata, by = "Sample_names")
```

```{r, include=TRUE, dpi = 500, fig.width = 8, fig.height = 5}
# Plot results
beta_fp %>% 
  ggplot(., aes(x = X1,  y = X2, fill = Treatment))+
  geom_point(shape = 21, size = 4)+
  # facet_wrap(.~Treatment, ncol = 3)+
  scale_fill_brewer("", palette = "Accent")+
  theme_bw()+
  labs(x = "PCoA 1", y = "PCoA 2")+
  theme(strip.text = element_text(size = 8))
```

## FlowFP

FlowFP does not seem to run in the current R and RStudio version. Example code
below in any case.

```{r, include=TRUE}
# fp_FP <-
#   flowFPModel(
#     fcs = FCS_resample(flowData_transformed_asinh_dn, 1e3),
#     param = param,
#     nRecursions = 8
#   )
```

## FlowEMMi

```{r, include=TRUE}
# fp_emmi <-
#   flowEMMi(
#     frame = flowData_transformed_asinh_dn[[1]],
#     ch1 = param[1],
#     ch2 = param[3],
#     sample_size = 1,
#     prior = FALSE,
#     separation = TRUE,
#     max_inits = 10,
#     use_log = FALSE,
#     alpha = .7,
#     img_format = "png",
#     foreground_maxsd = 5000,
#     start_cluster = 2,
#     end_cluster = 12
#   )
```

# Session info

```{r, include=TRUE}
sessionInfo()
```
